---
title: "BNCI Assignment 1"
output: html_notebook
---

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

# BNCI Assignment 1


## Preliminaries


### Loading data

```{r}
data_raw <- read.csv("winequality-white.csv", sep=";")
head(data_raw)
```


### Preprocessing

```{r}
data_normalized <- data_raw
data_normalized[ , 1:(ncol(data_normalized) - 1)] <- scale(data_raw[ , 1:(ncol(data_raw) - 1)]) # scaling
head(data_normalized)
```



### Inspecting data
```{r}
plot( residual.sugar ~ alcohol, data=data_raw, pch=19 )

plot( total.sulfur.dioxide ~ free.sulfur.dioxide , data=data_normalized, pch=19 )

plot( fixed.acidity ~ citric.acid , data=data_normalized, pch=19 )
```


## Reconstructing Paper

### Fitting SEM

```{r}
library(lavaan)
# the paper removed density!
# this is the one for white wine
model_paper <- sem( "
citric.acid ~ fixed.acidity + volatile.acidity
residual.sugar ~ citric.acid
alcohol ~ residual.sugar + volatile.acidity
free.sulfur.dioxide ~ fixed.acidity + volatile.acidity + alcohol + residual.sugar
chlorides ~ volatile.acidity + citric.acid + alcohol
free.sulfur.dioxide ~ fixed.acidity + volatile.acidity + residual.sugar + alcohol
total.sulfur.dioxide ~ free.sulfur.dioxide + alcohol
pH ~ fixed.acidity + residual.sugar + chlorides
sulphates ~ pH + free.sulfur.dioxide + total.sulfur.dioxide + chlorides
quality ~ pH + total.sulfur.dioxide + chlorides + citric.acid + volatile.acidity + alcohol
fixed.acidity ~~ 0*volatile.acidity
quality ~~ 0*sulphates
", # Do we need to specify empty connections here? How can I see which connections the model inferred?
data=data_normalized)
summary(model_paper)
```

### Converting and plotting with dagitty

```{r}
library( dagitty )
graph_paper <- lavaanToGraph(model_paper) 
graph_paper_plot <- lavaanToGraph(model_paper, digits = 4) 
plot(graph_paper_plot)
```

### Testing

```{r}
# localTests(graph_paper, data_normalized, type='cis.chisq') # I think this does not work as density was excluded?
CorM <- lavCor(data_normalized)
results <- localTests(graph_paper, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = TRUE)
results_ordered <- results[order(results$p.value),] #-abs(results$estimate),
results_ordered
```
```{r}
plotLocalTestResults( results )
```


## Our Model

### Initial iteration

```{r}
# Our proposed model

graph_white <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality
}')

#free.sulfur.dioxide -> volatile.acidity

plot(graph_white)
```

```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```


```{r}
plotLocalTestResults(results)
```
The model at the current state shows a huge misfit of density _||_ total.sulfur.dioxide. Looking into this revealed that it made sense. Total.sulfur.dioxide includes includes both the bound and free Sulfur dioxide (SO2). The main compound bound by sulfur dioxide is acetaldehyde which has a low density of 0.784 g/ mL. Therefore it makes sense that the bound compounds reduce the density of the wine. A direct edge can be established.


### Update 1 - total.sulfite.dioxide -> density

```{r}
graph_white_1 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality

total.sulfur.dioxide -> density

}')

plot(graph_white_1)
```

```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_1, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
```
As expected this resolves the worst misfitting.

The next misfitting implied coniditonal independence is alcohol _||_ total.sulfur.dioxide. Acetaldehyde plays a role here to as it is the compound that is transformed to alcohol by yeast during the fermentation process. When acetaldehyde is bound by sulphur dioxide though it cannot be transformed into alcohol. Therefore we introduce the connection between total.sulfur.dioxide and alchol.


### Update 2: total.sulfite.dioxide -> alcohol

```{r}
graph_white_2 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality


total.sulfur.dioxide -> density
total.sulfur.dioxide -> alcohol

}')

plot(graph_white_2)
```
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_2, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```
```{r}
plotLocalTestResults(results)
```
Again, a direct edge is guaranteed to resolve the misfit.

The next large problem is residual.sugar _||_ total.sulfur.dioxide. The process of fermentation seems like a fitting context here too. Here, sugar is consumed by yeast to produce alcohol. Sugar that is left when the fermentation process stopped is called residual sugar.

My idea to resolve this independence misfit is to find some latent representation of the fermentation process. I tried several explicit representations but they did not work out. When the SEM models where fit, the latent variable did not seem to express much of the variance. Then Johannes suggested to use implicit latent variables anyways. Those are much easier to fit and test as well.

Domain-wise, I would find that it makes more sense to fit free.sulfur.dioxide <-> residual.sugar, but that did not resolve the independence problem (after fitting this only finds a covariance of 6%)




### Update 3: Implicit latent variable Yeast

#### Just some investigation of the data
```{r}
# Plot 1: residual.sugar ~ total.sulfur.dioxide
plot(residual.sugar ~ total.sulfur.dioxide, data = data_normalized, pch = 19)
model1 <- lm(residual.sugar ~ total.sulfur.dioxide, data = data_normalized)
abline(model1, col = 2)

# Extract and print coefficient and p-value
summary1 <- summary(model1)
print(summary1)
coef1 <- summary1$coefficients["total.sulfur.dioxide", "Estimate"]
pval1 <- summary1$coefficients["total.sulfur.dioxide", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$total.sulfur.dioxide), 
     y = 0.9 * max(data_normalized$residual.sugar),
     labels = paste("Coeff =", round(coef1, 3), "\nP-value =", round(pval1, 3)), col = "blue")

# Plot 2: free.sulfur.dioxide ~ residual.sugar
plot(free.sulfur.dioxide ~ residual.sugar, data = data_normalized, pch = 19)
model2 <- lm(free.sulfur.dioxide ~ residual.sugar, data = data_normalized)
abline(model2, col = 2)

# Extract and print coefficient and p-value
summary2 <- summary(model2)
coef2 <- summary2$coefficients["residual.sugar", "Estimate"]
pval2 <- summary2$coefficients["residual.sugar", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$residual.sugar), 
     y = 0.9 * max(data_normalized$free.sulfur.dioxide),
     labels = paste("Coeff =", round(coef2, 3), "\nP-value =", round(pval2, 3)), col = "blue")

# Plot 3: free.sulfur.dioxide ~ total.sulfur.dioxide
plot(free.sulfur.dioxide ~ total.sulfur.dioxide, data = data_normalized, pch = 19)
model3 <- lm(free.sulfur.dioxide ~ total.sulfur.dioxide, data = data_normalized)
abline(model3, col = 2)

# Extract and print coefficient and p-value
summary3 <- summary(model3)
coef3 <- summary3$coefficients["total.sulfur.dioxide", "Estimate"]
pval3 <- summary3$coefficients["total.sulfur.dioxide", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$total.sulfur.dioxide), 
     y = 0.9 * max(data_normalized$free.sulfur.dioxide),
     labels = paste("Coeff =", round(coef3, 3), "\nP-value =", round(pval3, 3)), col = "blue")

```


#### Update of graph
```{r}
graph_white_3 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.673,0.581"]
chlorides [pos="0.233,0.498"]
citric.acid [pos="0.236,0.423"]
density [pos="0.494,0.464"]
fixed.acidity [pos="0.236,0.349"]
free.sulfur.dioxide [pos="0.228,0.200"]
pH [pos="0.667,0.348"]
quality [outcome,pos="0.494,0.199"]
residual.sugar [exposure,pos="0.237,0.585"]
sulphates [pos="0.658,0.125"]
total.sulfur.dioxide [pos="0.223,0.092"]
volatile.acidity [pos="0.232,0.276"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
total.sulfur.dioxide <-> residual.sugar [pos="0.140,0.380"]
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> alcohol
total.sulfur.dioxide -> density
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality
}
')

plot(graph_white_3)
```

#### SEM Fitting of implicit latent variable total.sulfur.dioxide <-> residual.sugar
```{r}
lvsem <- toString(graph_white_3,"lavaan") 
lvsem.fit <- sem(lvsem, data_normalized) 
summary(lvsem.fit)
```

#### Test of fit
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_3, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
```
The last remaining issue to be fixed is citric.acid _||_ fixed.acidity. Its not clear to me how you could explain this domain wise. The data set's paper says with fixed.acidity is the concentration of tartaric acid meant. But technically citric acid is also a fixed (non-volatile) acid and can be found naturally in wine with roughly 1/20 the amount of tartaric acid. This would support a direct edge between the two variables. But citric acid is also added manually as a metal binding agent and then its concentration would be independent of that of tartaric acid.
I decided to fit it as a latent variabel as the exact conditions are not known.


### Update 4: citric.acid <-> fixed.acidity

#### Investigation of citric.acid <-> fixed.acidity
```{r}
# Plot 1: residual.sugar ~ total.sulfur.dioxide
plot(citric.acid ~ fixed.acidity, data = data_normalized, pch = 19)
model1 <- lm(citric.acid ~ fixed.acidity, data = data_normalized)
abline(model1, col = 2)

# Extract and print coefficient and p-value
summary1 <- summary(model1)
print(summary1)
coef1 <- summary1$coefficients["fixed.acidity", "Estimate"]
pval1 <- summary1$coefficients["fixed.acidity", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$fixed.acidity), 
     y = 0.9 * max(data_normalized$citric.acid),
     labels = paste("Coeff =", round(coef1, 3), "\nP-value =", round(pval1, 3)), col = "blue")
```
#### Graph update
```{r}
graph_white_4 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.673,0.581"]
chlorides [pos="0.233,0.498"]
citric.acid [pos="0.236,0.423"]
density [pos="0.494,0.464"]
fixed.acidity [pos="0.236,0.349"]
free.sulfur.dioxide [pos="0.228,0.200"]
pH [pos="0.667,0.348"]
quality [outcome,pos="0.494,0.199"]
residual.sugar [exposure,pos="0.237,0.585"]
sulphates [pos="0.658,0.125"]
total.sulfur.dioxide [pos="0.223,0.092"]
volatile.acidity [pos="0.232,0.276"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
fixed.acidity <-> citric.acid [pos="0.170,0.400"]
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> alcohol
total.sulfur.dioxide -> density
total.sulfur.dioxide -> free.sulfur.dioxide
total.sulfur.dioxide <-> residual.sugar [pos="0.140,0.380"]
volatile.acidity -> pH
volatile.acidity -> quality
}
')

plot(graph_white_4)
```

#### SEM Fitting of implicit latent variable total.sulfur.dioxide <-> residual.sugar
```{r}
lvsem <- toString(graph_white_4,"lavaan") 
lvsem.fit <- sem(lvsem, data_normalized) 
summary(lvsem.fit)
```

#### Test of fit
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_4, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
```


### Calculating main causal effects
```{r}
graph_white_inference <- lavaanToGraph(lvsem_yeast_fit)
#impliedCovarianceMatrix(graph_white_inference)
```


Sanity check: Sulphates has no parents. Therefore a simple linear model should get similar results than what inference on our fitted graph finds.
```{r}
print(lm(quality ~ sulphates, data_latent))
print(impliedCovarianceMatrix( graph_white_inference)['sulphates','quality'])
```
seems good, its in the same magnitude.


## Omitted code

### Experiment 4.1: Latent modeling of Fermentation -> Failed
```{r}
graph_white_4 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality

residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality

total.sulfur.dioxide -> density
U.Fermentation [latent] U.Fermentation
U.Fermentation <- free.sulfur.dioxide
U.Fermentation -> residual.sugar
U.Fermentation -> alcohol
}')

# total.sulfur.dioxide -> alcohol
# residual.sugar -> alcohol

plot(graph_white_4)
```


#### SEM Fitting of explicit latent variable
```{r}
lvsem_fermentation <- "
alcohol ~ density
alcohol ~ pH
alcohol ~ quality
chlorides ~ alcohol
chlorides ~ density
chlorides ~ quality
citric.acid ~ density
citric.acid ~ pH
citric.acid ~ quality
density ~ quality
fixed.acidity ~ density
fixed.acidity ~ pH
fixed.acidity ~ quality
free.sulfur.dioxide ~ quality
pH ~ quality

residual.sugar ~ density
residual.sugar ~ quality
sulphates ~ quality
total.sulfur.dioxide ~ free.sulfur.dioxide
volatile.acidity ~ pH
volatile.acidity ~ quality

total.sulfur.dioxide ~ density
U.Fermentation =~ coeff1*free.sulfur.dioxide
U.Fermentation =~ coeff1*residual.sugar
U.Fermentation =~ coeff1*alcohol
"

# residual.sugar ~ alcohol

# define that coefficients are the same, restrict for identifiability
#U.Yeast =~ latent_coefficient*free.sulfur.dioxide
#U.Yeast =~ latent_coefficient*residual.sugar

lvsem_fermentation_fit <- sem(lvsem_fermentation, data_normalized)
#summary(lvsem_fermentation_fit)
```
#### Add latent variable to dataset
```{r}
data_latent <- data_normalized
latent_values <- predict(lvsem_fermentation_fit) 
data_latent$U.Fermentation <- as.numeric(latent_values)
head(data_latent)
```

```{r}
# Check variance of each variable
apply(data_latent, 2, var)
```


```{r}
#CorM <- lavCor(data_latent)
#results <- localTests(graph_white_4, sample.cov = CorM, sample.nobs = nrow(data_latent), abbreviate.names = FALSE)
#results_ordered <- results[order(-abs(results$estimate), results$p.value),]
#results_ordered

#Error: lavaan->lav_samplestats_icov():  
#   sample covariance matrix is not positive-definite

# The latent variable did not find reasonable relation.
```

### Experiment 4.2: Latent modeling of Fermentation -> Failed

```{r}
graph_white_4.2 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality

total.sulfur.dioxide -> density
total.sulfur.dioxide -> alcohol
U.Fermentation [latent] U.Fermentation
U.Fermentation -> free.sulfur.dioxide
U.Fermentation -> total.sulfur.dioxide
U.Fermentation -> residual.sugar
U.Fermentation -> alcohol
}')

plot(graph_white_4.2)
```


#### SEM Fitting of explicit latent variable
```{r}
lvsem_fermentation <- "
alcohol ~ density
alcohol ~ pH
alcohol ~ quality
chlorides ~ alcohol
chlorides ~ density
chlorides ~ quality
citric.acid ~ density
citric.acid ~ pH
citric.acid ~ quality
density ~ quality
fixed.acidity ~ density
fixed.acidity ~ pH
fixed.acidity ~ quality
free.sulfur.dioxide ~ quality
pH ~ quality
# residual.sugar ~ alcohol
residual.sugar ~ density
residual.sugar ~ quality
sulphates ~ quality
total.sulfur.dioxide ~ free.sulfur.dioxide
volatile.acidity ~ pH
volatile.acidity ~ quality

total.sulfur.dioxide ~ density
total.sulfur.dioxide ~ alcohol
U.Fermentation =~ coeff1*free.sulfur.dioxide
U.Fermentation =~ coeff1*total.sulfur.dioxide
U.Fermentation =~ coeff1*residual.sugar
U.Fermentation =~ coeff1*alcohol
"

lvsem_fermentation_fit <- sem(lvsem_fermentation, data_normalized)
summary(lvsem_fermentation_fit)
```


#### Add latent variable to dataset
```{r}
data_latent <- data_normalized
latent_values <- predict(lvsem_fermentation_fit) 
data_latent$U.Fermentation <- as.numeric(latent_values)
head(data_latent)
```

```{r}
# Check variance of each variable
apply(data_latent, 2, var)
```


```{r}
#CorM <- lavCor(data_latent)
#results <- localTests(graph_white_4, sample.cov = CorM, sample.nobs = nrow(data_latent), abbreviate.names = FALSE)
#results_ordered <- results[order(-abs(results$estimate), results$p.value),]
#results_ordered

#Error: lavaan->lav_samplestats_icov():  
#   sample covariance matrix is not positive-definite

# The latent variable did not find reasonable relation.
```





































