---
title: "BNCI Assignment 2"
output: html_notebook
---

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

# BNCI Assignment 2

## Preliminaries

### Loading data

```{r}
data_raw <- read.csv("winequality-white.csv", sep=";")
head(data_raw)
```


### Inspecting data

#### Plotting raw data
```{r}
# variable_names <- names(data_raw)
# 
# # Add a small value to zeros in citric.acid, such that log transforms can be applied
# # The rest of the variables seem all positive
# data_plot <- data_raw
# data_plot$citric.acid[data_plot$citric.acid == 0] <- 0.00000001
# 
# # Create a directory to save plots if needed
# plot_directory = "plots_A2"
#   
# if (!dir.exists(plot_directory)) {
#   dir.create(plot_directory)
# }
# 
# # Loop through variable pairs for plotting
# for (i in 1:(length(variable_names) - 1)) {
#   for (j in (i + 1):length(variable_names)) {
#     #if (i == 1 & j == 2){
#       var_x <- variable_names[i]
#       var_y <- variable_names[j]
#   
#       # Base plot
#       plot(data_plot[[var_x]], data_plot[[var_y]],
#            xlab = var_x, ylab = var_y,
#            pch = 19, main = paste(var_y, "vs", var_x))
#   
#       # Fit linear model
#       model <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
#       abline(model, col = 2)
#   
#       # Extract and display coefficient and p-value
#       summary_model <- summary(model)
#       coef <- summary_model$coefficients[2, "Estimate"]
#       pval <- summary_model$coefficients[2, "Pr(>|t|)"]
#       text(x = 0.7 * max(data_plot[[var_x]], na.rm = TRUE),
#            y = 0.9 * max(data_plot[[var_y]], na.rm = TRUE),
#            labels = paste("Coeff =", round(coef, 3), "\nP-value =", round(pval, 3)), col = "blue")
#   
#       # Save the plot
#       file_name <- paste0(plot_directory, "/", var_y, "_vs_", var_x, ".png")
#       dev.copy(png, file_name)
#       dev.off()
#   
#       # Log transformation plots
#       if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) { # Log works only for positive values
#         plot(log(data_plot[[var_x]]), log(data_plot[[var_y]]),
#              xlab = paste("log(", var_x, ")", sep = ""),
#              ylab = paste("log(", var_y, ")", sep = ""),
#              pch = 19, main = paste("log(", var_y, ") vs log(", var_x, ")"))
#   
#         # Fit linear model for log-transformed data
#         model_log <- lm(log(data_plot[[var_y]]) ~ log(data_plot[[var_x]]))
#         abline(model_log, col = 2)
#   
#         # Extract and display coefficient and p-value for log-transformed data
#         summary_model_log <- summary(model_log)
#         coef_log <- summary_model_log$coefficients[2, "Estimate"]
#         pval_log <- summary_model_log$coefficients[2, "Pr(>|t|)"]
#         text(x = 0.7 * max(log(data_plot[[var_x]]), na.rm = TRUE),
#              y = 0.9 * max(log(data_plot[[var_y]]), na.rm = TRUE),
#              labels = paste("Coeff =", round(coef_log, 3), "\nP-value =", round(pval_log, 3)), col = "blue")
#   
#         # Save the log-transformed plot
#         log_file_name <- paste0(plot_directory, "/", var_y, "_vs_log_", var_x, ".png")
#         dev.copy(png, log_file_name)
#         dev.off()
#     #}
#     }
#   }
# }

```

#### Residual plots
```{r}
# Create a directory to save residual plots if needed
# plot_directory = "res_plots_A2"
# if (!dir.exists(plot_directory)) {
#   dir.create(plot_directory)
# }
# 
# # Loop through variable pairs for residual plots
# for (i in 1:(length(variable_names) - 1)) {
#   for (j in (i + 1):length(variable_names)) {
#     var_x <- variable_names[i]
#     var_y <- variable_names[j]
# 
#     # Fit the linear model for raw data
#     model_raw <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
#     residuals_raw <- residuals(model_raw)
# 
#     # Plot residuals for raw data
#     plot(data_plot[[var_x]], residuals_raw,
#          xlab = var_x, ylab = paste("Residuals of", var_y),
#          pch = 19, main = paste("Residuals of", var_y, "vs", var_x))
# 
#     # Add a horizontal line at zero (representing the fitted model)
#     abline(h = 0, col = "red", lwd = 2, lty = 2)
# 
#     # Save the residual plot for raw data
#     file_name_raw <- paste0(plot_directory, "/raw_residuals_", var_y, "_vs_", var_x, ".png")
#     dev.copy(png, file_name_raw)
#     dev.off()
# 
#     # If both variables are positive, do the same for log-transformed data
#     if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
#       # Fit the linear model for log-transformed data
#       model_log <- lm(log(data_plot[[var_y]]) ~ log(data_plot[[var_x]]))
#       residuals_log <- residuals(model_log)
# 
#       # Plot residuals for log-transformed data
#       plot(log(data_plot[[var_x]]), residuals_log,
#            xlab = paste("log(", var_x, ")", sep = ""),
#            ylab = paste("Residuals of log(", var_y, ")", sep = ""),
#            pch = 19, main = paste("Residuals of log(", var_y, ") vs log(", var_x, ")"))
# 
#       # Add a horizontal line at zero (representing the fitted model)
#       abline(h = 0, col = "red", lwd = 2, lty = 2)
# 
#       # Save the residual plot for log-transformed data
#       file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
#       dev.copy(png, file_name_log)
#       dev.off()
#     }
#   }
# }
```




### Preprocessing

#### Applying log to alcohol seems useful
```{r}
hist(data_raw$chlorides)
hist(data_raw$alcohol)
var_x <- "chlorides"
var_y <- "alcohol"
plot_directory = "res_plots_A2"
data_plot = data_raw

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "alcohol"
var_y <- "chlorides"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "alcohol"
var_y <- "density"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "alcohol"
var_y <- "residual.sugar"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "residual.sugar"
var_y <- "alcohol"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}
```


#### Outlier removal
```{r}
# Removing residual sugar outlier
max_sugar <- max(data_raw$residual.sugar)
h <- hist(data_raw$residual.sugar, plot = FALSE)

plot(h, xlab = "Residual sugar concentration [g/L]", ylab = "Frequency",
     main = "Histogram of residual sugar in raw data", col = "blue")
abline(v = max_sugar, col = "red", lwd = 2, lty = 2)
legend("topright", 
       legend = paste("Outlier =", paste(round(max_sugar, 2), "g/L")),
       col = "red", 
       lty = 2, 
       lwd = 2)
print(max_sugar)
data_no_outlier <- data_raw[data_raw$residual.sugar != max_sugar,]
print(max(data_no_outlier$residual.sugar)) # outlier was removed
```

#### Plot outlier removal result
```{r}
var_x <- "residual.sugar"
var_y <- "alcohol"

# Plot raw data
data_plot = data_raw
# Fit the linear model for raw data
model_raw <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
residuals_raw <- residuals(model_raw)

# Plot residuals for raw data
plot(data_plot[[var_x]], residuals_raw, 
     xlab = var_x, ylab = paste("Residuals of", var_y), 
     pch = 19, main = paste("Residuals of", var_y, "vs", var_x))

# Add a horizontal line at zero (representing the fitted model)
abline(h = 0, col = "red", lwd = 2, lty = 2)


# Remove outlier and plot result
data_plot = data_no_outlier
# Fit the linear model for raw data
model_raw <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
residuals_raw <- residuals(model_raw)

# Plot residuals for raw data
plot(data_plot[[var_x]], residuals_raw, 
     xlab = var_x, ylab = paste("Residuals of", var_y), 
     pch = 19, main = paste("Residuals of", var_y, "vs", var_x))

# Add a horizontal line at zero (representing the fitted model)
abline(h = 0, col = "red", lwd = 2, lty = 2)

```

#### Combined preprocessing: Outlier removal, normalizing, log(alcohol)
```{r}
max_sugar <- max(data_raw$residual.sugar)
data_no_outlier <- data_raw[data_raw$residual.sugar != max_sugar,]
data_normalized <- data_no_outlier
data_normalized_incl_outlier <- data_raw

# Ordered factor for quality, such that lavaan applies polychoric relationship for quality
data_normalized$quality <- ordered(data_no_outlier$quality, levels = sort(unique(data_no_outlier$quality)))

# log transform of alcohol
data_normalized$alcohol <- log(data_normalized$alcohol)

# scaling for normalization
data_normalized[ , 1:(ncol(data_normalized) - 1)] <- scale(data_normalized[ , 1:(ncol(data_normalized) - 1)]) 
head(data_normalized)


# Including outlier for comparison
data_normalized_incl_outlier$quality <- ordered(data_normalized_incl_outlier$quality,
                                                levels = sort(unique(data_normalized_incl_outlier$quality))
                                                )
data_normalized_incl_outlier[ , 1:(ncol(data_normalized_incl_outlier) - 1)] <- scale(data_raw[ , 1:(ncol(data_raw) - 1)]) # scaling
```
```{r}
class(data_normalized["quality"])
hist(as.numeric(data_normalized$quality))
hist(data_no_outlier$alcohol)
hist(scale(data_no_outlier$alcohol))
hist(log(data_no_outlier$alcohol))
hist(scale(log(data_no_outlier$alcohol)))
hist(data_normalized$alcohol)
```



```{r}
# Sanity check: Ordered factor does not affect plotting
#data_test <- data_raw
#data_test$quality <- ordered(data_test$quality, levels = sort(unique(data_test$quality)))
data_test <- data_normalized
var_x = "fixed.acidity"
var_y = "quality"
plot(data_raw$fixed.acidity, data_raw$quality,
         xlab = var_x, ylab = var_y,
         pch = 19, main = paste(var_y, "vs", var_x))
plot(data_test$fixed.acidity, data_test$quality,
         xlab = var_x, ylab = var_y,
         pch = 19, main = paste(var_y, "vs", var_x))
```

## Our Model

### Initial iteration

```{r}
# Our proposed model
library(dagitty)

graph_white <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality
}')

#free.sulfur.dioxide -> volatile.acidity

plot(graph_white)
```

```{r}
library(lavaan)
CorM <- lavCor(data_normalized)
results <- localTests(graph_white, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```


```{r}
# , fig.width=16, fig.height=8
plotLocalTestResults(results)
#par = c("width = 12", "height=800", "res=150")
par(mar = c(5, 20, 4, 2)) #, cex.axis = 1.5, cex.lab = 2
plotLocalTestResults(results, n=10, auto.margin = FALSE)
```
The model at the current state shows a huge misfit of density _||_ total.sulfur.dioxide. Looking into this revealed that it made sense. Total.sulfur.dioxide includes includes both the bound and free Sulfur dioxide (SO2). The main compound bound by sulfur dioxide is acetaldehyde which has a low density of 0.784 g/ mL. Therefore it makes sense that the bound compounds reduce the density of the wine. A direct edge can be established.


### Update 1 - total.sulfite.dioxide -> density

```{r}
graph_white_1 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality

total.sulfur.dioxide -> density

}')

plot(graph_white_1)
```

```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_1, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results, n=10)
```
As expected this resolves the worst misfitting.

The next misfitting implied coniditonal independence is alcohol _||_ total.sulfur.dioxide. Acetaldehyde plays a role here to as it is the compound that is transformed to alcohol by yeast during the fermentation process. When acetaldehyde is bound by sulphur dioxide though it cannot be transformed into alcohol. Therefore we introduce the connection between total.sulfur.dioxide and alcohol.

```{r}
max_sugar <- max(data_raw$residual.sugar)
print(max_sugar)
data_no_outlier <- data_raw[data_raw$residual.sugar != max_sugar,]
data_normalized_no_outlier <- data_no_outlier

# Ordered factor for quality, such that lavaan applies polychoric relationship for quality
data_normalized_no_outlier$quality <- ordered(data_normalized$quality, levels = sort(unique(data_normalized_no_outlier$quality)))
data_normalized_no_outlier[ , 1:(ncol(data_normalized_no_outlier) - 1)] <- scale(data_raw[ , 1:(ncol(data_raw) - 1)]) # scaling
head(data_normalized_no_outlier)

CorM <- lavCor(data_normalized_no_outlier)
results <- localTests(graph_white, sample.cov = CorM, sample.nobs = nrow(data_normalized_no_outlier), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
plotLocalTestResults(results, n=10)
```


### Update 2: total.sulfite.dioxide -> alcohol

```{r}
graph_white_2 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality


total.sulfur.dioxide -> density
total.sulfur.dioxide -> alcohol

}')

plot(graph_white_2)
```
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_2, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```
```{r}
plotLocalTestResults(results)
```
Again, a direct edge is guaranteed to resolve the misfit.

The next large problem is residual.sugar _||_ total.sulfur.dioxide. The process of fermentation seems like a fitting context here too. Here, sugar is consumed by yeast to produce alcohol. Sugar that is left when the fermentation process stopped is called residual sugar.

My idea to resolve this independence misfit is to find some latent representation of the fermentation process. I tried several explicit representations but they did not work out. When the SEM models where fit, the latent variable did not seem to express much of the variance. Then Johannes suggested to use implicit latent variables anyways. Those are much easier to fit and test as well.

Domain-wise, I would find that it makes more sense to fit free.sulfur.dioxide <-> residual.sugar, but that did not resolve the independence problem (after fitting this only finds a covariance of 6%)




### Update 3: Implicit latent variableresidual.sugar _||_ total.sulfur.dioxide (e.g. Yeast)

#### Just some investigation of the data
```{r}
# Plot 1: residual.sugar ~ total.sulfur.dioxide
plot(residual.sugar ~ total.sulfur.dioxide, data = data_normalized, pch = 19)
model1 <- lm(residual.sugar ~ total.sulfur.dioxide, data = data_normalized)
abline(model1, col = 2)

# Extract and print coefficient and p-value
summary1 <- summary(model1)
print(summary1)
coef1 <- summary1$coefficients["total.sulfur.dioxide", "Estimate"]
pval1 <- summary1$coefficients["total.sulfur.dioxide", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$total.sulfur.dioxide), 
     y = 0.9 * max(data_normalized$residual.sugar),
     labels = paste("Coeff =", round(coef1, 3), "\nP-value =", round(pval1, 3)), col = "blue")

# Plot 2: free.sulfur.dioxide ~ residual.sugar
plot(free.sulfur.dioxide ~ residual.sugar, data = data_normalized, pch = 19)
model2 <- lm(free.sulfur.dioxide ~ residual.sugar, data = data_normalized)
abline(model2, col = 2)

# Extract and print coefficient and p-value
summary2 <- summary(model2)
coef2 <- summary2$coefficients["residual.sugar", "Estimate"]
pval2 <- summary2$coefficients["residual.sugar", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$residual.sugar), 
     y = 0.9 * max(data_normalized$free.sulfur.dioxide),
     labels = paste("Coeff =", round(coef2, 3), "\nP-value =", round(pval2, 3)), col = "blue")

# Plot 3: free.sulfur.dioxide ~ total.sulfur.dioxide
plot(free.sulfur.dioxide ~ total.sulfur.dioxide, data = data_normalized, pch = 19)
model3 <- lm(free.sulfur.dioxide ~ total.sulfur.dioxide, data = data_normalized)
abline(model3, col = 2)

# Extract and print coefficient and p-value
summary3 <- summary(model3)
coef3 <- summary3$coefficients["total.sulfur.dioxide", "Estimate"]
pval3 <- summary3$coefficients["total.sulfur.dioxide", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$total.sulfur.dioxide), 
     y = 0.9 * max(data_normalized$free.sulfur.dioxide),
     labels = paste("Coeff =", round(coef3, 3), "\nP-value =", round(pval3, 3)), col = "blue")

```


#### Update of graph
```{r}
graph_white_3 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.673,0.581"]
chlorides [pos="0.233,0.498"]
citric.acid [pos="0.236,0.423"]
density [pos="0.494,0.464"]
fixed.acidity [pos="0.236,0.349"]
free.sulfur.dioxide [pos="0.228,0.200"]
pH [pos="0.667,0.348"]
quality [outcome,pos="0.494,0.199"]
residual.sugar [exposure,pos="0.237,0.585"]
sulphates [pos="0.658,0.125"]
total.sulfur.dioxide [pos="0.223,0.092"]
volatile.acidity [pos="0.232,0.276"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
total.sulfur.dioxide <-> residual.sugar [pos="0.140,0.380"]
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> alcohol
total.sulfur.dioxide -> density
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality
}
')

plot(graph_white_3)
```

#### Test of fit of implicit latent variable total.sulfur.dioxide <-> residual.sugar
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_3, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
plotLocalTestResults(results, n=10)
```
The last remaining issue to be fixed is citric.acid _||_ fixed.acidity. Its not clear to me how you could explain this domain wise. The data set's paper says with fixed.acidity is the concentration of tartaric acid meant. But technically citric acid is also a fixed (non-volatile) acid and can be found naturally in wine with roughly 1/20 the amount of tartaric acid. This would support a direct edge between the two variables. But citric acid is also added manually as a metal binding agent and then its concentration would be independent of that of tartaric acid.
I decided to fit it as a latent variable as the exact conditions are not known.


### Update 4: implicit latent variable citric.acid <-> fixed.acidity

#### Investigation of citric.acid <-> fixed.acidity
```{r}
# Plot 1: residual.sugar ~ total.sulfur.dioxide
plot(citric.acid ~ fixed.acidity, data = data_normalized, pch = 19)
model1 <- lm(citric.acid ~ fixed.acidity, data = data_normalized)
abline(model1, col = 2)

# Extract and print coefficient and p-value
summary1 <- summary(model1)
print(summary1)
coef1 <- summary1$coefficients["fixed.acidity", "Estimate"]
pval1 <- summary1$coefficients["fixed.acidity", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$fixed.acidity), 
     y = 0.9 * max(data_normalized$citric.acid),
     labels = paste("Coeff =", round(coef1, 3), "\nP-value =", round(pval1, 3)), col = "blue")
```
#### Graph update
```{r}
graph_white_4 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.673,0.581"]
chlorides [pos="0.233,0.498"]
citric.acid [pos="0.236,0.423"]
density [pos="0.494,0.464"]
fixed.acidity [pos="0.236,0.349"]
free.sulfur.dioxide [pos="0.228,0.200"]
pH [pos="0.667,0.348"]
quality [outcome,pos="0.494,0.199"]
residual.sugar [exposure,pos="0.237,0.585"]
sulphates [pos="0.658,0.125"]
total.sulfur.dioxide [pos="0.223,0.092"]
volatile.acidity [pos="0.232,0.276"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
fixed.acidity <-> citric.acid [pos="0.170,0.400"]
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> alcohol
total.sulfur.dioxide -> density
total.sulfur.dioxide -> free.sulfur.dioxide
total.sulfur.dioxide <-> residual.sugar [pos="0.140,0.380"]
volatile.acidity -> pH
volatile.acidity -> quality
}
')

plot(graph_white_4)
```


#### Test of fit
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_4, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
plotLocalTestResults(results, n=10)
```

#### Fitting initial SEM
```{r}
# Note that there are several edges set 0 as they were implied by lavaan

lvsem <- "
density~alcohol
pH~alcohol
quality~alcohol
alcohol~chlorides
density~chlorides
quality~chlorides
density~citric.acid
pH~citric.acid
quality~citric.acid
citric.acid ~~ fixed.acidity
quality~density
density~fixed.acidity
pH~fixed.acidity
quality~fixed.acidity
quality~free.sulfur.dioxide
quality~pH
alcohol~residual.sugar
density~residual.sugar
quality~residual.sugar
total.sulfur.dioxide ~~ residual.sugar
quality~sulphates
alcohol~total.sulfur.dioxide
density~total.sulfur.dioxide
free.sulfur.dioxide~total.sulfur.dioxide
pH~volatile.acidity
quality~volatile.acidity

volatile.acidity ~~ 0*chlorides
volatile.acidity ~~ 0*sulphates
sulphates ~~ 0*chlorides
"

lvsem.fit <- sem(lvsem, data_normalized)
cg <- coordinates(graph_white_4)
fg <- lavaanToGraph(lvsem.fit, digits=2)
coordinates(fg) <- cg
plot(fg, show.coefficients=TRUE)
```
```{r}
# Extract all parameter estimates
params <- parameterEstimates(lvsem.fit)

# Filter for both regression (~) and bidirectional (~~) paths
edge_coeffs <- params[params$op %in% c("~", "~~"), ]

# Sort by magnitude in ascending order
edge_coeffs_sorted <- edge_coeffs[order(abs(edge_coeffs$est)), ]

# Display relevant columns (lhs, rhs, estimate)
edge_coeffs_sorted[, c("lhs", "rhs", "est")]
```
Pruning was done on in combinations of causal effects.

### Calculating main causal effects

#### General covariance matrix
```{r}
graph_white_inference <- lavaanToGraph(lvsem.fit)
#print(graph_white_inference)
imCovM <- impliedCovarianceMatrix(graph_white_inference)
```

Sanity check: Sulphates has no parents. Therefore a simple linear model should get similar results than what inference on our fitted graph finds.
```{r}
print(lm(quality ~ sulphates, data_normalized))
print(impliedCovarianceMatrix( graph_white_inference)['sulphates','quality'])
```
seems good, its in the same magnitude. But weird it has a different value


#### Function to create interventional graph
```{r}
# Function to create interventional graph
create_interventional_graph <- function(original_graph, variable_of_interest) {
  # Parse the original graph
  g <- dagitty(original_graph)
  
  # Extract node positions
  nodes <- names(g)
  coords <- coordinates(g)
  node_positions <- sapply(nodes, function(node) {
    sprintf('%s [pos="%.3f,%.3f"]', node, coords$x[node], coords$y[node])
  })
  
  # Extract edges and beta coefficients from the original graph string
  # Pattern to match: v ->|<-> w [beta="value"]
  beta_pattern <- "([a-zA-Z._]+) (->|<->) ([a-zA-Z._]+) \\[beta=\"([-0-9.e]+)\"\\]"
  betas <- stringr::str_match_all(original_graph, beta_pattern)[[1]]
  edges <- data.frame(v = betas[, 2], w = betas[, 4], e = betas[,3], beta = betas[, 5], stringsAsFactors = FALSE)
  
  # Filter out incoming edges to the variable of interest
  adjusted_edges <- edges[!(edges$e == "->" & edges$w == variable_of_interest), ]
  
  # Check for bidirectional edges and remove the ones involving the variable of interest
  bidirectional_edges <- edges[edges$e == "<->",]
  adjusted_bidirectional_edges <- bidirectional_edges[!(bidirectional_edges$v == variable_of_interest | bidirectional_edges$w == variable_of_interest), ]
  
  # Combine directed and filtered bidirectional edges
  adjusted_edges <- rbind(
    adjusted_edges[adjusted_edges$e == "->", ],
    adjusted_bidirectional_edges
  )
  
  # Create the adjusted DAG string
  interventional_graph <- sprintf("dag {\n%s\n}", 
                          paste(
                            node_positions,
                            paste0(
                              apply(adjusted_edges, 1, function(row) {
                                if (row["e"] == "->") {
                                  sprintf("%s -> %s [beta=\"%s\"]", row["v"], row["w"], row["beta"])
                                } else {
                                  sprintf("%s <-> %s [beta=\"%s\"]", row["v"], row["w"], row["beta"])
                                }
                              }), 
                              collapse = "\n"
                            ), 
                            collapse = "\n"
                          ))
  
  return(dagitty(interventional_graph))
}
```

#### Testing the interventional graph function
```{r}
# Generate interventional graph for "fixed.acidity"
intervention_graph <- create_interventional_graph(fg, "density")
cat(fg) #original graph
cat(intervention_graph)
plot(intervention_graph, show.coefficients = TRUE)
```
#### Calculating causal effects
```{r}
nodes <- names(fg)
# Initialize a data frame to store the results
causal_effects <- data.frame(
  Variable = character(),  # To store variable names
  CausalEffect = numeric() # To store causal effects
)

# Iterate over the nodes, skipping the "quality" variable
for (node in nodes) {
  if (node == "quality") next  # Skip the "quality" variable
  
  # Create the intervention graph
  intervention_graph <- create_interventional_graph(fg, node)
  
  # Calculate the causal effect of the node on quality
  causal_effect <- impliedCovarianceMatrix(intervention_graph)[node, 'quality']
  
  # Append the result to the data frame
  causal_effects <- rbind(causal_effects, data.frame(Variable = node, CausalEffect = causal_effect))
}

# Print the final data frame
#print(causal_effects)
ce_ordered <- causal_effects[order(-abs(causal_effects$CausalEffect)),]
print(ce_ordered)
```
Therefore our first research question is answered. Listed are the causal effects sorted by size of influence. 
- Alcohol seems in line with what other researcher found. As alcohol was always in the top rankings of importance
- density is surprising
- volatile acidity also seems logical. Acetic acid makes wine taste like vinegar. Its an indicator for spoiled wine. A strong negative effect seems logical though
- A negative effect of total.sulfur.dioxide also seems reasonable as sulfuric compounds are normally perceived as not pleasant (e.g. rotting eggs smell is due to sulfur compounds)
- Surprising are citric.acid and residual.sugar and their low scores as these play a large role for the taste of wine. But I guess if the wine is sweet or acidic does not automatically indicate if its good or bad quality but has more characterizing property.

Using polychoric correlations changed some of this:
- now density has overly large contribution. This does not make much sense. 
- free.sulfur.dioxide is listed with a positive causal effect which is weird. See the interpretation of total.sulfur dioxide
- pH has a very low coefficient in the causal effect as well as the SEM model. As the acids in the wine directly influence the quality (in taste in acidity) it makes sense to exclude the relation pH -> quality. It then ends up as a collider in the network which also is logical as it is a mathematical measure of the total acids presents.
- I would restrain to remove other edges as these are thought to give taste characteristics to the wine. It is informative to know if their effect is small or large.

### SEM Refitting without pH
```{r}
# Now pH is also set to 0

lvsem <- "
density~alcohol
pH~alcohol
quality~alcohol
alcohol~chlorides
density~chlorides
quality~chlorides
density~citric.acid
pH~citric.acid
quality~citric.acid
citric.acid ~~ fixed.acidity
quality~density
density~fixed.acidity
pH~fixed.acidity
quality~fixed.acidity
quality~free.sulfur.dioxide

pH ~~ 0*quality

alcohol~residual.sugar
density~residual.sugar
quality~residual.sugar
total.sulfur.dioxide ~~ residual.sugar
quality~sulphates
alcohol~total.sulfur.dioxide
density~total.sulfur.dioxide
free.sulfur.dioxide~total.sulfur.dioxide
pH~volatile.acidity
quality~volatile.acidity

volatile.acidity ~~ 0*chlorides
volatile.acidity ~~ 0*sulphates
sulphates ~~ 0*chlorides
"

lvsem.fit <- sem(lvsem, data_normalized)
summary(lvsem.fit)
cg <- coordinates(graph_white_4)
fg <- lavaanToGraph(lvsem.fit, digits=2)
coordinates(fg) <- cg
plot(fg, show.coefficients=TRUE)
```
#### Test of fit
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_4, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
par(mar = c(5, 20, 4, 2))
plotLocalTestResults(results, n=10, auto.margin=FALSE)
```

#### Model coefficients
```{r}
# Extract all parameter estimates
params <- parameterEstimates(lvsem.fit)

# Filter for both regression (~) and bidirectional (~~) paths
edge_coeffs <- params[params$op %in% c("~", "~~"), ]

# Sort by magnitude in ascending order
edge_coeffs_sorted <- edge_coeffs[order(abs(edge_coeffs$est)), ]

# Display relevant columns (lhs, rhs, estimate)
edge_coeffs_sorted[, c("lhs", "rhs", "est")]
```
#### Recalculating Causal Effects
```{r}
nodes <- names(fg)
# Initialize a data frame to store the results
causal_effects <- data.frame(
  Variable = character(),  # To store variable names
  CausalEffect = numeric() # To store causal effects
)

# Iterate over the nodes, skipping the "quality" variable
for (node in nodes) {
  if (node == "quality") next  # Skip the "quality" variable
  
  # Create the intervention graph
  intervention_graph <- create_interventional_graph(fg, node)
  
  # Calculate the causal effect of the node on quality
  causal_effect <- impliedCovarianceMatrix(intervention_graph)[node, 'quality']
  
  # Append the result to the data frame
  causal_effects <- rbind(causal_effects, data.frame(Variable = node, CausalEffect = causal_effect))
}

# Print the final data frame
#print(causal_effects)
ce_ordered <- causal_effects[order(-abs(causal_effects$CausalEffect)),]
ce_ordered$CausalEffect <- round(ce_ordered$CausalEffect, 3)
print(ce_ordered)
```
#### Interpretation of causal effects

```{r}
# Initialize a data frame for denormalized causal effects
denormalized_effects <- data.frame(
  Variable = character(),           # Parent variables
  NormalizedEffect = numeric(),   # Normalized causal effect
  parent_sd_raw = numeric(),
  DenormalizedEffect = numeric()  # Denormalized causal effect
)

# Iterate over the variables (assuming causal_effects is your calculated data frame)

for (i in 1:nrow(causal_effects)) {
  parent <- causal_effects$Variable[i]
  #print(parent)
  normalized_effect <- causal_effects$CausalEffect[i]
  
  # Get the standard deviation of the parent variable
  #print(c(data_raw[parent]))
  parent_sd <- sd(data_raw[[parent]])
  #print(parent_sd)
  
  # Get the standard deviation of the quality variable
  child_sd <- sd(as.numeric(data_raw$quality))  # Treat quality as numeric for SD
  
  # Denormalize the effect
  denormalized_effect <- normalized_effect * parent_sd / child_sd
  
  # Append to the data frame
  denormalized_effects <- rbind(denormalized_effects, 
                                data.frame(
                                  Variable = parent,
                                  NormalizedEffect = normalized_effect,
                                  parent_sd_raw = parent_sd,
                                  DenormalizedEffect = denormalized_effect
                                ))
}

# Round results to three decimals
denormalized_effects$NormalizedEffect <- round(denormalized_effects$NormalizedEffect, 3)
denormalized_effects$parent_sd_raw <- round(denormalized_effects$parent_sd_raw, 3)
denormalized_effects$DenormalizedEffect <- round(denormalized_effects$DenormalizedEffect, 3)

# Sort by the absolute value of the normalized effect
denormalized_effects <- denormalized_effects[order(-abs(denormalized_effects$NormalizedEffect)), ]

# Print the denormalized causal effects
print(denormalized_effects)
```

### Structure Learning
```{r}
variable_names <- 
print(variable_names)
```

```{r}
library(bnlearn)

whitelist <- data.frame(
  from = c('alcohol'), 
  to = c('quality')
)

quality_strings <- c('quality', 'quality', 'quality', 'quality', 'quality',
                     'quality', 'quality', 'quality', 'quality', 'quality',
                     'quality', 'quality')
to_strings <- names(data_raw) # all variables
blacklist <- data.frame(
  from = quality_strings,
  to = to_strings
)

print(whitelist)

fit <- pc.stable(data_normalized, whitelist = whitelist, blacklist = blacklist)
print(fit)
```
```{r}
plot(fit)
```



























