---
title: "BNCI Assignment 1"
output: html_notebook
---

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

# BNCI Assignment 1


## Preliminaries


### Loading data

```{r}
data_raw <- read.csv("winequality-white.csv", sep=";")
head(data_raw)
```


### Preprocessing

```{r}
data_normalized <- data_raw
data_normalized[ , 1:(ncol(data_normalized) - 1)] <- scale(data_raw[ , 1:(ncol(data_raw) - 1)]) # scaling
head(data_normalized)
```


## Reconstructing Paper

### Fitting SEM

```{r}
library(lavaan)
# the paper removed density!
# this is the one for white wine
model_paper <- sem( "
citric.acid ~ fixed.acidity + volatile.acidity
residual.sugar ~ citric.acid
alcohol ~ residual.sugar + volatile.acidity
free.sulfur.dioxide ~ fixed.acidity + volatile.acidity + alcohol + residual.sugar
chlorides ~ volatile.acidity + citric.acid + alcohol
free.sulfur.dioxide ~ fixed.acidity + volatile.acidity + residual.sugar + alcohol
total.sulfur.dioxide ~ free.sulfur.dioxide + alcohol
pH ~ fixed.acidity + residual.sugar + chlorides
sulphates ~ pH + free.sulfur.dioxide + total.sulfur.dioxide + chlorides
quality ~ pH + total.sulfur.dioxide + chlorides + citric.acid + volatile.acidity + alcohol
fixed.acidity ~~ 0*volatile.acidity
quality ~~ 0*sulphates
", # Do we need to specify empty connections here? How can I see which connections the model inferred?
data=data_normalized)
summary(model_paper)
```

### Converting and plotting with dagitty

```{r}
library( dagitty )
graph_paper <- lavaanToGraph(model_paper) 
graph_paper_plot <- lavaanToGraph(model_paper, digits = 4) 
plot(graph_paper_plot)
```

### Testing

```{r}
# localTests(graph_paper, data_normalized, type='cis.chisq') # I think this does not work as density was excluded?
CorM <- lavCor(data_normalized)
results <- localTests(graph_paper, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = TRUE)
plotLocalTestResults( results )
```

## Our Model

### Graph definition

```{r}
# Our proposed model

graph_white <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
free.sulfur.dioxide -> volatile.acidity
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality
}')

plot(graph_white)
```
### Testing
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results
```


```{r}
plotLocalTestResults(results)
```

# Remove this header
```{r}
x <- "Remove this code block."
x
```

