---
title: "BNCI Assignment 2"
output: html_notebook
---

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

# BNCI Assignment 2

## Preliminaries

### Loading data

```{r}
data_raw <- read.csv("winequality-white.csv", sep=";")
head(data_raw)
```


### Inspecting data

#### Plotting raw data
```{r}
# variable_names <- names(data_raw)
# 
# # Add a small value to zeros in citric.acid, such that log transforms can be applied
# # The rest of the variables seem all positive
# data_plot <- data_raw
# data_plot$citric.acid[data_plot$citric.acid == 0] <- 0.00000001
# 
# # Create a directory to save plots if needed
# plot_directory = "plots_A2"
# 
# if (!dir.exists(plot_directory)) {
#   dir.create(plot_directory)
# }
# 
# # Loop through variable pairs for plotting
# for (i in 1:(length(variable_names) - 1)) {
#   for (j in (i + 1):length(variable_names)) {
#     #if (i == 1 & j == 2){
#       var_x <- variable_names[i]
#       var_y <- variable_names[j]
# 
#       # Base plot
#       plot(data_plot[[var_x]], data_plot[[var_y]],
#            xlab = var_x, ylab = var_y,
#            pch = 19, main = paste(var_y, "vs", var_x))
# 
#       # Fit linear model
#       model <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
#       abline(model, col = 2)
# 
#       # Extract and display coefficient and p-value
#       summary_model <- summary(model)
#       coef <- summary_model$coefficients[2, "Estimate"]
#       pval <- summary_model$coefficients[2, "Pr(>|t|)"]
#       text(x = 0.7 * max(data_plot[[var_x]], na.rm = TRUE),
#            y = 0.9 * max(data_plot[[var_y]], na.rm = TRUE),
#            labels = paste("Coeff =", round(coef, 3), "\nP-value =", round(pval, 3)), col = "blue")
# 
#       # Save the plot
#       file_name <- paste0(plot_directory, "/", var_y, "_vs_", var_x, ".png")
#       dev.copy(png, file_name)
#       dev.off()
# 
#       # Log transformation plots
#       if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) { # Log works only for positive values
#         plot(log(data_plot[[var_x]]), log(data_plot[[var_y]]),
#              xlab = paste("log(", var_x, ")", sep = ""),
#              ylab = paste("log(", var_y, ")", sep = ""),
#              pch = 19, main = paste("log(", var_y, ") vs log(", var_x, ")"))
# 
#         # Fit linear model for log-transformed data
#         model_log <- lm(log(data_plot[[var_y]]) ~ log(data_plot[[var_x]]))
#         abline(model_log, col = 2)
# 
#         # Extract and display coefficient and p-value for log-transformed data
#         summary_model_log <- summary(model_log)
#         coef_log <- summary_model_log$coefficients[2, "Estimate"]
#         pval_log <- summary_model_log$coefficients[2, "Pr(>|t|)"]
#         text(x = 0.7 * max(log(data_plot[[var_x]]), na.rm = TRUE),
#              y = 0.9 * max(log(data_plot[[var_y]]), na.rm = TRUE),
#              labels = paste("Coeff =", round(coef_log, 3), "\nP-value =", round(pval_log, 3)), col = "blue")
# 
#         # Save the log-transformed plot
#         log_file_name <- paste0(plot_directory, "/", var_y, "_vs_log_", var_x, ".png")
#         dev.copy(png, log_file_name)
#         dev.off()
#     #}
#     }
#   }
# }

```

#### Residual plots
```{r}
# #Create a directory to save residual plots if needed
# plot_directory = "res_plots_A2"
# if (!dir.exists(plot_directory)) {
#   dir.create(plot_directory)
# }
# 
# # Loop through variable pairs for residual plots
# for (i in 1:(length(variable_names) - 1)) {
#   for (j in (i + 1):length(variable_names)) {
#     var_x <- variable_names[i]
#     var_y <- variable_names[j]
# 
#     # Fit the linear model for raw data
#     model_raw <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
#     residuals_raw <- residuals(model_raw)
# 
#     # Plot residuals for raw data
#     plot(data_plot[[var_x]], residuals_raw,
#          xlab = var_x, ylab = paste("Residuals of", var_y),
#          pch = 19, main = paste("Residuals of", var_y, "vs", var_x))
# 
#     # Add a horizontal line at zero (representing the fitted model)
#     abline(h = 0, col = "red", lwd = 2, lty = 2)
# 
#     # Save the residual plot for raw data
#     file_name_raw <- paste0(plot_directory, "/raw_residuals_", var_y, "_vs_", var_x, ".png")
#     dev.copy(png, file_name_raw)
#     dev.off()
# 
#     # If both variables are positive, do the same for log-transformed data
#     if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
#       # Fit the linear model for log-transformed data
#       model_log <- lm(log(data_plot[[var_y]]) ~ log(data_plot[[var_x]]))
#       residuals_log <- residuals(model_log)
# 
#       # Plot residuals for log-transformed data
#       plot(log(data_plot[[var_x]]), residuals_log,
#            xlab = paste("log(", var_x, ")", sep = ""),
#            ylab = paste("Residuals of log(", var_y, ")", sep = ""),
#            pch = 19, main = paste("Residuals of log(", var_y, ") vs log(", var_x, ")"))
# 
#       # Add a horizontal line at zero (representing the fitted model)
#       abline(h = 0, col = "red", lwd = 2, lty = 2)
# 
#       # Save the residual plot for log-transformed data
#       file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
#       dev.copy(png, file_name_log)
#       dev.off()
#     }
#   }
# }
```




### Preprocessing

#### Applying log to alcohol seems useful
```{r}
hist(data_raw$chlorides)
hist(data_raw$alcohol)

var_x <- "chlorides"
var_y <- "alcohol"
plot_directory = "res_plots_A2"
data_plot = data_raw

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "alcohol"
var_y <- "chlorides"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "alcohol"
var_y <- "density"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "alcohol"
var_y <- "residual.sugar"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}

var_x <- "residual.sugar"
var_y <- "alcohol"

if (all(data_plot[[var_x]] > 0) && all(data_plot[[var_y]] > 0)) {
      # Fit the linear model for log-transformed data
      model_log <- lm(data_plot[[var_y]] ~ log(data_plot[[var_x]]))
      residuals_log <- residuals(model_log)

      # Plot residuals for log-transformed data
      plot(log(data_plot[[var_x]]), residuals_log,
           xlab = paste("log(", var_x, ")", sep = ""),
           ylab = paste("Residuals of ", var_y, "", sep = ""),
           pch = 19, main = paste("Residuals of ", var_y, " vs log(", var_x, ")"))

      # Add a horizontal line at zero (representing the fitted model)
      abline(h = 0, col = "red", lwd = 2, lty = 2)

      # Save the residual plot for log-transformed data
      file_name_log <- paste0(plot_directory, "/log_residuals_", var_y, "_vs_", var_x, ".png")
      dev.copy(png, file_name_log)
      dev.off()
}
```


#### Outlier removal
```{r}
# Removing residual sugar outlier
max_sugar <- max(data_raw$residual.sugar)
h <- hist(data_raw$residual.sugar, plot = FALSE)

plot(h, xlab = "Residual sugar concentration [g/L]", ylab = "Frequency",
     main = "Histogram of residual sugar in raw data", col = "blue")
abline(v = max_sugar, col = "red", lwd = 2, lty = 2)
legend("topright", 
       legend = paste("Outlier =", paste(round(max_sugar, 2), "g/L")),
       col = "red", 
       lty = 2, 
       lwd = 2)
print(max_sugar)
data_no_outlier <- data_raw[data_raw$residual.sugar != max_sugar,]
print(max(data_no_outlier$residual.sugar)) # outlier was removed
```

#### Plot outlier removal result
```{r}
var_x <- "residual.sugar"
var_y <- "alcohol"

# Plot raw data
data_plot = data_raw
# Fit the linear model for raw data
model_raw <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
residuals_raw <- residuals(model_raw)

# Plot residuals for raw data
plot(data_plot[[var_x]], residuals_raw, 
     xlab = var_x, ylab = paste("Residuals of", var_y), 
     pch = 19, main = paste("Residuals of", var_y, "vs", var_x))

# Add a horizontal line at zero (representing the fitted model)
abline(h = 0, col = "red", lwd = 2, lty = 2)


# Remove outlier and plot result
data_plot = data_no_outlier
# Fit the linear model for raw data
model_raw <- lm(data_plot[[var_y]] ~ data_plot[[var_x]])
residuals_raw <- residuals(model_raw)

# Plot residuals for raw data
plot(data_plot[[var_x]], residuals_raw, 
     xlab = var_x, ylab = paste("Residuals of", var_y), 
     pch = 19, main = paste("Residuals of", var_y, "vs", var_x))

# Add a horizontal line at zero (representing the fitted model)
abline(h = 0, col = "red", lwd = 2, lty = 2)

```

#### Combined preprocessing: Outlier removal, normalizing, log(alcohol)
```{r}
max_sugar <- max(data_raw$residual.sugar)
data_no_outlier <- data_raw[data_raw$residual.sugar != max_sugar,]
data_normalized <- data_no_outlier
data_normalized_incl_outlier <- data_raw

# Ordered factor for quality, such that lavaan applies polychoric relationship for quality
data_normalized$quality <- ordered(data_no_outlier$quality, levels = sort(unique(data_no_outlier$quality)))

# log transform of alcohol
data_normalized$alcohol <- log(data_normalized$alcohol)

# scaling for normalization
data_normalized[ , 1:(ncol(data_normalized) - 1)] <- scale(data_normalized[ , 1:(ncol(data_normalized) - 1)]) 
head(data_normalized)


# Including outlier for comparison
data_normalized_incl_outlier$quality <- ordered(data_normalized_incl_outlier$quality,
                                                levels = sort(unique(data_normalized_incl_outlier$quality))
                                                )
data_normalized_incl_outlier[ , 1:(ncol(data_normalized_incl_outlier) - 1)] <- scale(data_raw[ , 1:(ncol(data_raw) - 1)]) # scaling
```
```{r}
# Sanity check preprocessing
class(data_normalized["quality"])
hist(as.numeric(data_normalized$quality))
hist(data_no_outlier$alcohol)
hist(scale(data_no_outlier$alcohol))
hist(log(data_no_outlier$alcohol))
hist(scale(log(data_no_outlier$alcohol)))
hist(data_normalized$alcohol)
```



```{r}
# Sanity check: Ordered factor does not affect plotting
#data_test <- data_raw
#data_test$quality <- ordered(data_test$quality, levels = sort(unique(data_test$quality)))
data_test <- data_normalized
var_x = "fixed.acidity"
var_y = "quality"
plot(data_raw$fixed.acidity, data_raw$quality,
         xlab = var_x, ylab = var_y,
         pch = 19, main = paste(var_y, "vs", var_x))
plot(data_test$fixed.acidity, data_test$quality,
         xlab = var_x, ylab = var_y,
         pch = 19, main = paste(var_y, "vs", var_x))
```

## Adjustments

### Initial iteration

```{r}
# Our proposed model
library(dagitty)

graph_white <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality
}')

#free.sulfur.dioxide -> volatile.acidity

plot(graph_white)
```

```{r}
library(lavaan)
CorM <- lavCor(data_normalized)
results <- localTests(graph_white, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```


```{r}
# , fig.width=16, fig.height=8
plotLocalTestResults(results)
#par = c("width = 12", "height=800", "res=150")
par(mar = c(5, 20, 4, 2)) #, cex.axis = 1.5, cex.lab = 2
plotLocalTestResults(results, n=10, auto.margin = FALSE)
```
The model at the current state shows a huge misfit of density _||_ total.sulfur.dioxide. Looking into this revealed that it made sense. Total.sulfur.dioxide includes includes both the bound and free Sulfur dioxide (SO2). The main compound bound by sulfur dioxide is acetaldehyde which has a low density of 0.784 g/ mL. Therefore it makes sense that the bound compounds reduce the density of the wine. A direct edge can be established.


### Update 1 - total.sulfite.dioxide -> density

```{r}
graph_white_1 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality

total.sulfur.dioxide -> density

}')

plot(graph_white_1)
```

```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_1, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results, n=10)
```
As expected this resolves the worst misfitting.

The next misfitting implied coniditonal independence is alcohol _||_ total.sulfur.dioxide. Acetaldehyde plays a role here to as it is the compound that is transformed to alcohol by yeast during the fermentation process. When acetaldehyde is bound by sulphur dioxide though it cannot be transformed into alcohol. Therefore we introduce the connection between total.sulfur.dioxide and alcohol.


### Update 2: total.sulfite.dioxide -> alcohol

```{r}
graph_white_2 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.713,0.893"]
chlorides [pos="0.165,0.786"]
citric.acid [pos="0.169,0.671"]
density [pos="0.433,0.687"]
fixed.acidity [pos="0.167,0.538"]
free.sulfur.dioxide [pos="0.172,0.281"]
pH [pos="0.719,0.515"]
quality [outcome,pos="0.433,0.378"]
residual.sugar [exposure,pos="0.164,0.895"]
sulphates [pos="0.699,0.194"]
total.sulfur.dioxide [pos="0.173,0.156"]
volatile.acidity [pos="0.171,0.434"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality


total.sulfur.dioxide -> density
total.sulfur.dioxide -> alcohol

}')

plot(graph_white_2)
```

```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_2, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
```
Again, a direct edge is guaranteed to resolve the misfit.

The next large problem is residual.sugar _||_ total.sulfur.dioxide. The process of fermentation seems like a fitting context here too. Here, sugar is consumed by yeast to produce alcohol. Sugar that is left when the fermentation process stopped is called residual sugar.

My idea to resolve this independence misfit is to find some latent representation of the fermentation process. I tried several explicit representations but they did not work out. When the SEM models where fit, the latent variable did not seem to express much of the variance. Then Johannes suggested to use implicit latent variables anyways. Those are much easier to fit and test as well.

Domain-wise, I would find that it makes more sense to fit free.sulfur.dioxide <-> residual.sugar, but that did not resolve the independence problem (after fitting this only finds a covariance of 6%)


### Update 3: Implicit latent variableresidual.sugar _||_ total.sulfur.dioxide (e.g. Yeast)

#### Just some investigation of the data
```{r}
# Plot 1: residual.sugar ~ total.sulfur.dioxide
plot(residual.sugar ~ total.sulfur.dioxide, data = data_normalized, pch = 19)
model1 <- lm(residual.sugar ~ total.sulfur.dioxide, data = data_normalized)
abline(model1, col = 2)

# Extract and print coefficient and p-value
summary1 <- summary(model1)
print(summary1)
coef1 <- summary1$coefficients["total.sulfur.dioxide", "Estimate"]
pval1 <- summary1$coefficients["total.sulfur.dioxide", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$total.sulfur.dioxide), 
     y = 0.9 * max(data_normalized$residual.sugar),
     labels = paste("Coeff =", round(coef1, 3), "\nP-value =", round(pval1, 3)), col = "blue")

# Plot 2: free.sulfur.dioxide ~ residual.sugar
plot(free.sulfur.dioxide ~ residual.sugar, data = data_normalized, pch = 19)
model2 <- lm(free.sulfur.dioxide ~ residual.sugar, data = data_normalized)
abline(model2, col = 2)

# Extract and print coefficient and p-value
summary2 <- summary(model2)
coef2 <- summary2$coefficients["residual.sugar", "Estimate"]
pval2 <- summary2$coefficients["residual.sugar", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$residual.sugar), 
     y = 0.9 * max(data_normalized$free.sulfur.dioxide),
     labels = paste("Coeff =", round(coef2, 3), "\nP-value =", round(pval2, 3)), col = "blue")

# Plot 3: free.sulfur.dioxide ~ total.sulfur.dioxide
plot(free.sulfur.dioxide ~ total.sulfur.dioxide, data = data_normalized, pch = 19)
model3 <- lm(free.sulfur.dioxide ~ total.sulfur.dioxide, data = data_normalized)
abline(model3, col = 2)

# Extract and print coefficient and p-value
summary3 <- summary(model3)
coef3 <- summary3$coefficients["total.sulfur.dioxide", "Estimate"]
pval3 <- summary3$coefficients["total.sulfur.dioxide", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$total.sulfur.dioxide), 
     y = 0.9 * max(data_normalized$free.sulfur.dioxide),
     labels = paste("Coeff =", round(coef3, 3), "\nP-value =", round(pval3, 3)), col = "blue")

```


#### Update of graph
```{r}
graph_white_3 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.673,0.581"]
chlorides [pos="0.233,0.498"]
citric.acid [pos="0.236,0.423"]
density [pos="0.494,0.464"]
fixed.acidity [pos="0.236,0.349"]
free.sulfur.dioxide [pos="0.228,0.200"]
pH [pos="0.667,0.348"]
quality [outcome,pos="0.494,0.199"]
residual.sugar [exposure,pos="0.237,0.585"]
sulphates [pos="0.658,0.125"]
total.sulfur.dioxide [pos="0.223,0.092"]
volatile.acidity [pos="0.232,0.276"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> free.sulfur.dioxide
volatile.acidity -> pH
volatile.acidity -> quality

total.sulfur.dioxide -> alcohol
total.sulfur.dioxide -> density
total.sulfur.dioxide <-> residual.sugar [pos="0.140,0.380"]
}
')

plot(graph_white_3)
```

#### Test of fit of implicit latent variable total.sulfur.dioxide <-> residual.sugar
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_3, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
plotLocalTestResults(results, n=10)
```
The last remaining issue to be fixed is citric.acid _||_ fixed.acidity. Its not clear to me how you could explain this domain wise. The data set's paper says with fixed.acidity is the concentration of tartaric acid meant. But technically citric acid is also a fixed (non-volatile) acid and can be found naturally in wine with roughly 1/20 the amount of tartaric acid. This would support a direct edge between the two variables. But citric acid is also added manually as a metal binding agent and then its concentration would be independent of that of tartaric acid.
I decided to fit it as a latent variable as the exact conditions are not known.


### Update 4: implicit latent variable citric.acid <-> fixed.acidity

#### Investigation of citric.acid <-> fixed.acidity
```{r}
# Plot 1: residual.sugar ~ total.sulfur.dioxide
plot(citric.acid ~ fixed.acidity, data = data_normalized, pch = 19)
model1 <- lm(citric.acid ~ fixed.acidity, data = data_normalized)
abline(model1, col = 2)

# Extract and print coefficient and p-value
summary1 <- summary(model1)
print(summary1)
coef1 <- summary1$coefficients["fixed.acidity", "Estimate"]
pval1 <- summary1$coefficients["fixed.acidity", "Pr(>|t|)"]
text(x = 0.7 * max(data_normalized$fixed.acidity), 
     y = 0.9 * max(data_normalized$citric.acid),
     labels = paste("Coeff =", round(coef1, 3), "\nP-value =", round(pval1, 3)), col = "blue")
```
#### Graph update
```{r}
graph_white_4 <- dagitty('dag {
bb="0,0,1,1"
alcohol [pos="0.673,0.581"]
chlorides [pos="0.233,0.498"]
citric.acid [pos="0.236,0.423"]
density [pos="0.494,0.464"]
fixed.acidity [pos="0.236,0.349"]
free.sulfur.dioxide [pos="0.228,0.200"]
pH [pos="0.667,0.348"]
quality [outcome,pos="0.494,0.199"]
residual.sugar [exposure,pos="0.237,0.585"]
sulphates [pos="0.658,0.125"]
total.sulfur.dioxide [pos="0.223,0.092"]
volatile.acidity [pos="0.232,0.276"]
alcohol -> density
alcohol -> pH
alcohol -> quality
chlorides -> alcohol
chlorides -> density
chlorides -> quality
citric.acid -> density
citric.acid -> pH
citric.acid -> quality
density -> quality
fixed.acidity -> density
fixed.acidity -> pH
fixed.acidity -> quality
fixed.acidity <-> citric.acid [pos="0.170,0.400"]
free.sulfur.dioxide -> quality
pH -> quality
residual.sugar -> alcohol
residual.sugar -> density
residual.sugar -> quality
sulphates -> quality
total.sulfur.dioxide -> alcohol
total.sulfur.dioxide -> density
total.sulfur.dioxide -> free.sulfur.dioxide
total.sulfur.dioxide <-> residual.sugar [pos="0.140,0.380"]
volatile.acidity -> pH
volatile.acidity -> quality
}
')

plot(graph_white_4)
```


#### Test of fit
```{r}
CorM <- lavCor(data_normalized)
results <- localTests(graph_white_4, sample.cov = CorM, sample.nobs = nrow(data_normalized), abbreviate.names = FALSE)
results_ordered <- results[order(-abs(results$estimate), results$p.value),]
results_ordered
```

```{r}
plotLocalTestResults(results)
plotLocalTestResults(results, n=10)
```


## Fitting adjusted SEM

Note that the connection pH -> quality is pruned here as layed out in assignment 1.
I did not want to keep the causal inference section I created in the first assignment as it is the main focus of the second assignment.

```{r}
# Note that there are several edges set 0 as they were implied by lavaan

lvsem <- "
density~alcohol
pH~alcohol
quality~alcohol
alcohol~chlorides
density~chlorides
quality~chlorides
density~citric.acid
pH~citric.acid
quality~citric.acid
citric.acid ~~ fixed.acidity
quality~density
density~fixed.acidity
pH~fixed.acidity
quality~fixed.acidity
quality~free.sulfur.dioxide

pH ~~ 0*quality

alcohol~residual.sugar
density~residual.sugar
quality~residual.sugar
total.sulfur.dioxide ~~ residual.sugar
quality~sulphates
alcohol~total.sulfur.dioxide
density~total.sulfur.dioxide
free.sulfur.dioxide~total.sulfur.dioxide
pH~volatile.acidity
quality~volatile.acidity

volatile.acidity ~~ 0*chlorides
volatile.acidity ~~ 0*sulphates
sulphates ~~ 0*chlorides
"

lvsem.fit <- sem(lvsem, data_normalized, ordered = c("quality"))
cg <- coordinates(graph_white_4)
fg <- lavaanToGraph(lvsem.fit, digits=2)
coordinates(fg) <- cg
plot(fg, show.coefficients=TRUE)
```
```{r}
# Sanity check: Lavaan did incorporate the polychoric correlations for the quality variable. As can be seen in the summary, it assigned thresholds for the different quality levels.
summary(lvsem.fit)
```

```{r}
# Extract all parameter estimates
params <- parameterEstimates(lvsem.fit)

# Filter for both regression (~) and bidirectional (~~) paths
edge_coeffs <- params[params$op %in% c("~", "~~"), ]

# Sort by magnitude in ascending order
edge_coeffs_sorted <- edge_coeffs[order(abs(edge_coeffs$est)), ]

# Display relevant columns (lhs, rhs, estimate)
edge_coeffs_sorted[, c("lhs", "rhs", "est")]
```
Pruning was done on in combinations of causal effects.

## Causal Inference

### Covariate Adjustment

##### General covariance matrix
```{r}
graph_white_inference <- lavaanToGraph(lvsem.fit)
#print(graph_white_inference)
imCovM <- impliedCovarianceMatrix(graph_white_inference)
```

Sanity check: Sulphates has no parents. Therefore a simple linear model should get similar results than what inference on our fitted graph finds.
```{r}
print(lm(quality ~ sulphates, data_normalized))
print(impliedCovarianceMatrix( graph_white_inference)['sulphates','quality'])
```
seems good, its in the same magnitude. But weird it has a different value
Edit: Probably as lm() is not suited for the ordered factor as the warning states.Better suited would be something like logistic regression which can handle (ordered) categorical data. Lavaan fits the ordered factor via polychoric correlations though.

#### Via Interventional graphs
I decided to switch over to the regression adjustment as our treatment variables are not binary. E.g. the alcohol level can be changed with different percentage levels. Its not including alcohol and not including alcohol

#### Via adjustment sets

```{r}
alcohol_adjustment <- adjustmentSets(graph_white_inference, exposure = "alcohol", outcome = "quality")
print(alcohol_adjustment)
```
```{r}
# Test, normalization is difficult to interpret together with log transform of alcohol.
# Run regression again without the normalization

# Unadjusted estimate
#mdl <- glm(quality ~ alcohol, 
#            data = data_normalized, 
#            family = "binomial")
library(MASS)
# I'm using data_normalized here, simply because quality is saved as the ordered factor in that dataframe.
mdl <- polr(data_normalized$quality ~ log(data_no_outlier$alcohol), 
            method = "logistic")  # or "probit" for a probit link
summary(mdl)


# Interpret coefficients
# coeff = change in logs(odds_ratio) per unit change in log(alcohol)
# exp(coeff) = change in odds_ratio per unit change in log(alcohol)
# log(1.01) = represents the change in log scale for a 1% increase in alcohol
# exp(coef*log(1.01)) = change in odds_ratio upon a 1% increase in alcohol

exp(coef(mdl))
exp(coef(mdl)*log(1.10))
stde <- summary(mdl)$coefficients[1, "Std. Error"]
#print(stde)
exp(stde*log(1.10))
```


```{r}
# Adjusted estimate
#mdl <- glm(quality ~ alcohol+chlorides+residual.sugar+total.sulfur.dioxide, 
#            data = data_normalized, 
#            family = "binomial")
mdl <- polr(data_normalized$quality ~ log(alcohol)+chlorides+residual.sugar+total.sulfur.dioxide, 
            data = data_no_outlier, 
            method = "logistic")  # or "probit" for a probit link
normalized_coefficients <- coef(mdl)
summary(mdl)

# Interpret coefficients
print("Exponentiated coefficients")
exp(coef(mdl))
print("Alcohol 10% change")
exp(coef(mdl)*log(1.10))["log(alcohol)"]
print("Alcohol 10% change -> Standard errors")
stde <- summary(mdl)$coefficients[1, "Std. Error"]
print(stde)
exp(stde*log(1.10))
```

### Propensity Scores - Adjustment
```{r}
#Another sanity check about preprocessing
hist(data_raw$alcohol)
hist(log(data_raw$alcohol), breaks=20)
#hist(scale(data_raw$alcohol), breaks=20)
hist(scale(log(data_raw$alcohol)), breaks=20)
hist(data_normalized$alcohol, breaks=20)
```


```{r}
data = data_no_outlier
mdl <- lm( log(alcohol) ~ chlorides + residual.sugar + total.sulfur.dioxide, data )
pscores <- dnorm(mdl$residuals, sd=sigma(mdl))
hist(mdl$residuals)
```
Assumption for propensity scores to work is that this linear regression for the scores is correct. That is the case when the residuals are normally distributed.

```{r}
# Omitted, we dont use normalized data anymore for better interpretability.

#bins <- length(sort(unique(data_raw$quality)))
#print(sum(data_raw$quality == 3))
#hist(data_raw$quality)
#quality_normalized <- scale(data_no_outlier$quality)
#hist(quality_normalized)
# hist(data_normalized$quality) does not work as stored as ordered factor here


```


For the propensity score regression we do not treat quality as an ordered factor but as continous, which should be decent.
Simply for the reason that the course did not treat the case of an ordinal outcome variable and we did not find a reasonably easy attempt to account for this.

```{r}
# Unadjusted effect

mdl <- lm(data$quality ~ log(data$alcohol))

summary(mdl)
coef(mdl)*log(1.10)
stde <- summary(mdl)$coefficients[2, "Std. Error"]
#print(stde)
stde*log(1.10)

# Omitted: Denormalized effect:

# Normalized effect
#normalized_effect <- coef(mdl)[2]
#print(normalized_effect)
# denormalized effect = normalized effect ∗ variable std/quality std
#parent_sd <- sd(data$alcohol)
#child_sd <- sd(data$quality)
#denormalized_effect <- normalized_effect * parent_sd / child_sd
#print(denormalized_effect)
```

```{r}
# Same procedure for adjusted effect

mdl <- lm(data$quality ~ log(data$alcohol) + pscores)

summary(mdl)
coef(mdl)*log(1.10)
stde <- summary(mdl)$coefficients[2:3, "Std. Error"]
stde*log(1.10)

# Omitted: 
# Normalized effect
#normalized_effect <- coef(mdl)[2]
#print(normalized_effect)
# Denormalized effect
#parent_sd <- sd(data_no_outlier$alcohol)
#child_sd <- sd(data_no_outlier$quality)
#denormalized_effect <- normalized_effect * parent_sd / child_sd
#print(denormalized_effect)
```

## Omitted Code

### Left out normalisation from logistic regression. Otherwise interpretation is too difficult
```{r}
# Unadjusted estimate
#mdl <- glm(quality ~ alcohol, 
#            data = data_normalized, 
#            family = "binomial")
library(MASS)
mdl <- polr(quality ~ alcohol, 
            data = data_normalized, 
            method = "logistic")  # or "probit" for a probit link
normalized_coefficients <- coef(mdl)
summary(mdl)

# Interpret coefficients
exp(coef(mdl))  # Get odds ratios for each predictor
```

```{r}
# Converting normalized data back for interpretation

# Initialize a dataframe to store results
results_df <- data.frame(
  Predictor = names(normalized_coefficients),
  Normalized_Coefficient = as.numeric(normalized_coefficients),
  Normalized_Odds_Ratio = as.numeric(exp(normalized_coefficients)),
  Std_Dev_Original = NA,  # Placeholder for standard deviations
  Denormalized_Coefficient = NA,  # Placeholder for denormalized coefficients
  Denormalized_Odds_Ratio = NA  # Placeholder for odds ratios
)

# Loop through each predictor
for (predictor in results_df$Predictor) {
  # Compute the original standard deviation of the predictor
  sd_original <- sd(data_no_outlier[[predictor]])
  
  # Denormalize the coefficient
  denormalized_coeff <- normalized_coefficients[predictor] * sd_original
  
  # Compute the odds ratio
  odds_ratio <- exp(denormalized_coeff)
  
  # Store values in the dataframe
  results_df[results_df$Predictor == predictor, "Std_Dev_Original"] <- sd_original
  results_df[results_df$Predictor == predictor, "Denormalized_Coefficient"] <- denormalized_coeff
  results_df[results_df$Predictor == predictor, "Denormalized_Odds_Ratio"] <- odds_ratio
}

# Display the results
print(results_df)
```

```{r}
# Adjusted estimate
#mdl <- glm(quality ~ alcohol+chlorides+residual.sugar+total.sulfur.dioxide, 
#            data = data_normalized, 
#            family = "binomial")
mdl <- polr(quality ~ alcohol+chlorides+residual.sugar+total.sulfur.dioxide, 
            data = data_normalized, 
            method = "logistic")  # or "probit" for a probit link
normalized_coefficients <- coef(mdl)
summary(mdl)

# Interpret coefficients
exp(coef(mdl))  # Get odds ratios for each predictor
```

```{r}
# Converting normalized data back for interpretation

# Initialize a dataframe to store results
results_df <- data.frame(
  Predictor = names(normalized_coefficients),
  Normalized_Coefficient = as.numeric(normalized_coefficients),
  Normalized_Odds_Ratio = as.numeric(exp(normalized_coefficients)),
  Std_Dev_Original = NA,  # Placeholder for standard deviations
  Denormalized_Coefficient = NA,  # Placeholder for denormalized coefficients
  Denormalized_Odds_Ratio = NA  # Placeholder for odds ratios
)

# Loop through each predictor
for (predictor in results_df$Predictor) {
  # Compute the original standard deviation of the predictor
  sd_original <- sd(data_no_outlier[[predictor]])
  
  # Denormalize the coefficient
  denormalized_coeff <- normalized_coefficients[predictor] * sd_original
  
  # Compute the odds ratio
  odds_ratio <- exp(denormalized_coeff)
  
  # Store values in the dataframe
  results_df[results_df$Predictor == predictor, "Std_Dev_Original"] <- sd_original
  results_df[results_df$Predictor == predictor, "Denormalized_Coefficient"] <- denormalized_coeff
  results_df[results_df$Predictor == predictor, "Denormalized_Odds_Ratio"] <- odds_ratio
}

# Display the results
print(results_df)
```

### as Team of 2: Structure Learning
```{r}
#variable_names <- 
#print(variable_names)
```

```{r}
library(bnlearn)

whitelist <- data.frame(
  from = c('alcohol'), 
  to = c('quality')
)

quality_strings <- c('quality', 'quality', 'quality', 'quality', 'quality',
                     'quality', 'quality', 'quality', 'quality', 'quality',
                     'quality', 'quality')
to_strings <- names(data_raw) # all variables
blacklist <- data.frame(
  from = quality_strings,
  to = to_strings
)

print(whitelist)

fit <- pc.stable(data_normalized, whitelist = whitelist, blacklist = blacklist)
print(fit)
```

```{r}
plot(fit)
```



























